---
title: Numerical Optimum Allocation
output:
  html_document: 
    theme: readable
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, out.width = "100%", fig.align = "center", cache = FALSE, dev = ifelse(knitr::is_html_output(), "png", "pdf"), comment = "")
```

```{r packages, echo = FALSE}
library(tidyverse)
library(kableExtra)
```

```{r utilities, echo = FALSE}
source("../../utilities.R")
```

In class we discussed formulas for computing the sample sizes for optimum allocation for a stratified random sampling design. These formulas can be derived *analytically* using calculus and what are called Lagrange multipliers. Another approach is to solve the optimum allocation problem *numerically* by computer. This has a couple of advantages. One is that it lets us avoid having to do mathematical derivations which can be difficult, time-consuming, and error-prone. Of course, we do not need to re-do the derivations for optimum allocation for the cases we discussed in class since we already have the equations, but we would if we were to (a) consider a different estimator with a different variance or (b) a different cost function. Another advantage is that the numerical approach allows us to find a solution where the sample sizes are constrained such that $2 \le n_j \le N_j$.

Optimum allocation is a constrained optimization problem. There are several R packages available for solving such problems. Here I will demonstrate using the package **Rsolnp**. First we need to install and load the package.
```{r, eval = FALSE}
install.packages("Rsolnp")
library(Rsolnp)
```
```{r, echo = FALSE}
library(Rsolnp)
```
Note that you only need to install the package via `install.packages` once per installation of R on a given computer. This downloads the package and installs it. But the `library` command needs to be used each time you restart R to make its contents available (namely the `solnp` function we will be using below).

Next we need to program functions functions to compute the variance of an estimator, the cost of the survey (if we want to have a fixed cost), and/or the bound on the error of estimation (if we want to have a fixed bound). Recall that the variance of $\hat\mu$ is
$$
  V(\hat\mu) = \frac{1}{N^2}\sum_{j=1}^L N_j^2\left(1 - \frac{n_j}{N_j}\right)\frac{\sigma_j^2}{n_j}.
$$
We can program a function `vf` to compute the variance of $\hat\mu$. 
```{r}
vf <- function(nj, Nj, sj, c0, cj) {
  1 / sum(Nj)^2 * sum(Nj^2 * (1 - nj / Nj) * sj^2 / nj)
}
```
Here `nj`, `Nj`, and `sj` represent the $n_j$, $N_j$, and $\sigma_j$ for the $L$ strata, respectively. But what might be confusing is that each of these variables represent the quantities for *all* strata --- i.e., the `nj` represents $n_1, n_2, \dots, n_L$. These are what are sometimes called vectors or (one-dimensional) arrays. R supports "vectorized" calculations where we can effectively program calculations on sets of values at once. Something else that may be confusing is that I have included as arguments to the function `c0` and `cj` which represent what we called $c_0$ and $c_j$ for the cost of the survey. The variance is not a function of either of these quantities, but they are included here because the function we will use for the constrained optimization requires that every function passed to it has the same arguments.

Recall that we defined the cost of a survey as
$$
  C = c_0 + \sum_{j=1}^L n_j c_j.
$$
We can program this as follows.
```{r}
cf <- function(nj, Nj, sj, c0, cj) {
  c0 + sum(nj * cj)
}
```
Finally the bound on the error of estimation is 
$$
B = 2\sqrt{V(\hat\mu)}
$$
which we can program as follows using the `vf` function to compute $V(\hat\mu)$.
```{r}
bf <- function(nj, Nj, sj, c0, cj) {
  2 * sqrt(vf(nj, Nj, sj, c0, cj))
}
```
Now we are ready for the constrained optimization to compute optimum allocations. Here I will replicate the examples from lecture for the sword fern survey using the following data.
```{r, echo = FALSE}
d <- data.frame(Stratum = 1:2, Region = c("Forest","Prairie"), Ni = c(30,87), ni = c(8,5),
  mi = c(287,11.3), si = c(149.1,16.8))
d$Region <- as.character(d$Region)
d <- rbind(d, c("","",117,13,"",""))
names(d)[3:6] <- c("$N_j$","$n_j$","$\\bar{y}_j$","$s_j$")
ktbl(d)
```
First consider the problem of minimizing the cost of the survey for a fixed bound on the error of estimation of $B$ = 20 $g/m^2$. We can find the solution as follows using the function `solnp` from the **Rsolnp** package.
```{r}
tmp <- solnp(pars = c(5,5), fun = cf, eqfun = bf, eqB = 20,
  Nj = c(30,87), sj = c(149.1,16.8), cj = c(4,1), c0 = 20,
  LB = c(2,2), UB = c(30,87))
tmp$par
```
This may require a bit of explanation. The first argument `par` are "starting values" for the two sample sizes. These need not be the actual sample sizes for the optimum allocation. They are just "rough guesses" to get the algorithm started. The argument `fun` is the function for what we are minimizing, which is the cost of the survey. The argument `eqfun` is the function for what we want to fix (an equality constraint), which is the bound on the error of estimation. The argument `eqB` is the value we want to fix (here the bound on the error of estimation). The arguments `Nj`, `sj`, `cj`, and `c0` pass some necessary information to these functions. And finally `LB` and `UB` are the lower-bound and upper-bound to the sample sizes for the optimum allocation. These are not needed here since the sample sizes for the optimum allocation are both between two and the strata sizes, but this is how those constraints would be included. Notice that the solution shown using `tmp$par` is within rounding error of what was found in lecture.

Now consider the problem of minimizing the bound on the error of estimation for a fixed total cost of $C$ = 100 and an overhead cost of $c_0$ = 20. This can be done as follows.
```{r}
tmp <- solnp(pars = c(5,5), fun = bf, eqfun = cf, eqB = 100,
  Nj = c(30,87), sj = c(149.1,16.8), cj = c(4,1), c0 = 20,
  LB = c(2,2), UB = c(30,87))
tmp$par
```
Note that all that has changed is the function defining what we are minimizing (`bf`), the function for what we are fixing (`cf`), and the value we are fixing (`eqB`). Again, the solution is within rounding error of what we found in lecture. 




