---
title: "Abundance Estimation"
output:
  html_document: 
    theme: readable
header-includes:
  - \usepackage{array}
  - \usepackage{booktabs}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, echo = FALSE, out.width = "100%", fig.align = "center", cache = FALSE, comment = "", dev = ifelse(knitr::is_html_output(), "png", "pdf"))
```

```{r packages, message = FALSE}
library(tidyverse)
library(kableExtra)
```

```{r utilities}
source("../../utilities.R")
```

`r ifelse(knitr::is_html_output(), paste("You can also download a [PDF](hw-abundance.pdf) copy of this document.", sep = ""), "")`

```{r}
d <- trtools::larkspur
N <- 150
n <- 20
p <- 0.3
ybar <- mean(d$plants)
xbar <- mean(d$area)
taux <- 5897
yinf <- round(sum(d$plants) * 0.3)
tauy <- 4400
```

### Estimation of the Abundance of Larkspur

Conservation researchers conducted a survey to estimate the abundance of a species of "larkspur" (a flowering plant of the genus *Delphinium*) in an irregularly-shaped region. To survey the larkspur they divided the region into `r N` one-meter wide transects from north to south. But because of the irregular shape of the region the transects were not all of equal area. The total area of the region (and hence all of the transects) was `r taux` square meters. The researchers selected a sample of `r n` transects using simple random sampling. The mean number of plants in these sampled transects was `r ybar` larkspur, and the mean area of these sampled transects was `r xbar` square meters. Note that we can view this design as a one-stage cluster sampling design where the transects are the clusters and the larkspur are the elements.[^srscluster] For your calculations below note that $n\bar{y} = \sum_{i \in \mathcal{S}} y_i$ and $n\bar{x} = \sum_{i \in \mathcal{S}} x_i$, so you can find those sums for your calculations from the number of sampled transects and the mean of the variable. You may also find it useful to note that $y_i = m_i$ if $y_i$ is the number of larkspur in the $i$-th transect. 

[^srscluster]: Recall that when clusters are selected using simple random sampling, a one-stage cluster sampling design can also be viewed as a simple random sampling design if we view the clusters as elements and the cluster totals --- i.e., $y_i = \sum_{j=1}^{m_i} y_{ij}$ as the target variable.

1. Using the "unbiased" estimator (i.e., the estimator that does not use an auxiliary variable), confirm that the estimate of the number of larkspur in the region is `r rnd(N * ybar)` larkspur. Using the "ratio" estimator with transect area as the auxiliary variable, confirm that the estimate of the number of larkspur in the region is `r rnd(taux * ybar/xbar)` larkspur.

0. Now suppose that the researchers are interested in estimating the number of larkspur in the region that are infected with a fungus. The researchers find that the total number of *infected* larkspur in the sampled transects is `r yinf` plants. Confirm that using the "unbiased" estimator with no auxiliary variable that the estimated number of infected larkspur is `r N/n * yinf` plants. Also confirm that using the ratio estimator with the number of larkspur in the transects as an auxiliary variable that the estimated number of infected larkspur is `r rnd(tauy * yinf / sum(d$plants))` (assume that we somehow know that the total number of larkspur in the region is `r tauy`). Finally confirm that using the ratio estimator with the transect area as an auxiliary variable that the estimated number of infected larkspur is `r rnd(taux * yinf / sum(d$area))`.

### We Just Can't Leaf it Alone

```{r}
set.seed(123)
d <- trtools::leafarea
d$weight <- d$area / 50 + rnorm(nrow(d), 0, d$area/400)
n <- 20
N <- 744
p <- 0.4
xbar <- round(mean(d$weight), 2)
taux <- round(sum(N, mean(d$weight), sd(d$weight)/sqrt(n)),2)
```

Recall once again the example from an earlier homework where the researchers had a population of `r N` leaves and used leaf weight as an auxiliary variable to estimate mean leaf area. They obtained a simple random sample of `r n` leaves and found that the mean weight of the leaves in that sample was `r xbar` grams. They also know that the total weight of all the leaves is `r taux` grams. 

1. Suppose the researchers *do not know* the number of leaves in the population (i.e., they do not know that $N$ = `r N`). Using a ratio estimator with leaf weight as the auxiliary variable, confirm that the estimated number of leaves in the population is `r rnd(taux*(1/xbar))` leaves. 

0. Now suppose the researchers want to estimate the number of leaves in the population that are infected with a particular fungus. In their sample of `r n` leaves they find that `r p*n` of the leaves are infected with the fungus. There are a couple of ways they could estimate the number of infected leaves in the population. One is that if they *do not know* $N$ they could use a *ratio estimator* with weight as an auxiliary variable. The other is that if they *know* $N$ they could estimate the number of infected leaves *without* the auxiliary variable. Confirm that the estimate from the first estimator is `r rnd(taux*p/xbar)` infected leaves, and that the estimate from the second estimator is `r rnd(N*p)` infected leaves. 

```{r}
N <- 500
n <- 100
y <- n * 0.3 
x <- n * 0.25
tx <- N * 0.2
```

### Foot Lice

Researchers at the National Institutes of Hobbit Health (NIHH) want to estimate the number of Hobbits in a village of `r N` Hobbits that have foot lice. Accurate diagnosis of the presence of foot lice is an expensive time-consuming process, so the researchers used a survey with a simple random sampling design to obtain a sample of `r n` Hobbits. In this sample `r y` Hobbits were found to have foot lice. The researchers had also mailed to each Hobbit in the village a home test kit where they could self-diagnose the presence of foot lice. Hobbits that detected the presence of foot lice using the home test were asked to report the positive result to the NIHH. The home test is not completely accurate, and also some Hobbits might not report their positive test result, and some Hobbits might never use their home test.[^sensspec] But it could still be used as an auxiliary variable. A total of `r tx` Hobbits from the village reported a positive test result to the NIHH. In the sample, `r x` of the Hobbits had reported a positive test result using the home test. Use the information above to confirm that the estimate of the number of Hobbits infected with foot lice using the expansion estimator is `r N*y/n` Hobbits, and the estimate using the ratio estimator (with the reporting of a positive test result as the auxiliary variable) is `r tx * y/x` Hobbits.

[^sensspec]: Here the sensitivity and specificity is not just a function of the diagnostic qualities of the home test. They also depend on the cooperation (or lack thereof) of the Hobbit!

### Jelly Beans

```{r, echo = FALSE}
t <- 50
n <- 100
s <- 16
N.lin <- t*n/s
N.chp <- (t + 1) * (n + 1) / (s + 1) - 1
N.mle <- floor(t*n/s)
```
  
Suppose I have a jar of jelly beans. I remove a handful of `r t` jelly beans, carefully poke a hole through each jelly bean using a needle, and then put those jelly beans back in the jar. I then shake the jar and remove a handful of `r n` jelly beans. In this sample of jelly beans I notice that `r s` of the jelly beans have holes. Confirm that the estimates of the number of jelly beans in the jar is `r N.lin` jelly beans using the *Lincoln-Petersen* estimator, `r N.chp` jelly beans using the *Chapman* estimator, and `r N.mle` jelly beans using the *maximum likelihood* estimator (based on the hypergeometric model).

### Hypergeometric Hobbits

```{r hypergeometric, echo = FALSE}
N <- 260:700
t <- 100
n <- 200
s <- 40
l <- dhyper(s, t, N-t, n)
```

Researchers used a mark-recapture study to estimate the population of Hobbits in the East Farthing. After marking an initial sample of Hobbits, they obtained a second sample of `r n` Hobbits, and found that `r s` of these were marked. The figure below shows the likelihood function for the parameter $N$ based on the hypergeometric model. What is the *approximate* estimate of $\hat{N}$ based on the maximum likelihood estimator?

```{r likelihood, fig.width = 8, echo = FALSE}
d <- data.frame(N = N, l = l)
p <- ggplot(d, aes(x = N, y = l)) + theme_minimal()
p <- p + geom_point(size = 0.1) + labs(x = "N", y = "Likelihood")
plot(p)
```

### Overlapping Lists

```{r cr, echo = FALSE}
m1 <- 1504 
n <- 1535
m2 <- 794
lp <- function(m1, n, m2) {
  return(m1 * n / m2)
}
```

Mark-recapture methodology is useful with humans where reliable records are unavailable. Typically these studies do not literally mark people. Instead they use lists of individuals. An early real example of this methodology was a study from 1949 that estimated the number of births in an area near Calcutta, India. The researchers had access to two lists of births --- lists *A* and *B* (one of the lists was from the Maternity and Child Welfare Department, and the other was complied by the All-India Institute of Hygiene and Public Health). Neither list contained all births. List *A* recorded `r m1` births, and list *B* recorded `r n` births. When the two lists were compared (births were recorded by name so common records could be identified), it was observed that there were `r m2` births that were on *both* lists. The data can be summarized in a table tabulating the number of actual births that were present or absent from each list.
```{r}
d <- data.frame(first = c("included","excluded","Total"),
  included = c(m2,n-m2,n),
  excluded = c(m1-m2,"?","?"),
  Total = c(m1,"?","$N$"))
names(d)[1] <- c("List A")
ktbl(d, align = "lccc") %>% add_header_above(c(" " = 1, "List B" = 2, " " = 1)) %>% column_spec(1, bold = TRUE)
```
Confirm that the estimate of the number of births using the Lincoln-Petersen estimator is about `r round(lp(m1, n, m2))` births.

### Widgets in a Warehouse

```{r, echo = FALSE}
N <- 1000
n <- 3
taux <- 5500
y <- c(20,25,30)
x <- c(5,4,6)
tauy <- (mean(y)/mean(x)) * taux
```

Recall the examples from class where "marked" elements might simply be a subset of known elements (e.g., the known shipwrecks in an area or the known cases of employees reporting sexual harassment). In one-stage cluster sampling designs the number of "marked" units per cluster can then be used as an auxiliary variable in ratio estimators. Consider a survey to estimate the number of widgets in a warehouse. The widgets are contained in a total of `r N` boxes that each contains several widgets. There are several types of widgets in the warehouse and it is known that there are exactly `r taux` Type IV widgets in the warehouse (the number of widgets of other types is unknown). A sample of `r n` boxes is selected using simple random sampling. The first box contains `r y[1]` widgets, of which `r x[1]` are Type IV widgets. The second box contains `r y[2]` widgets, of which `r x[2]` are Type IV widgets. The third box contains `r y[3]` widgets, of which `r x[3]` are Type IV widgets. There are a couple of ways we could estimate the number of widgets (of any type) in the warehouse. One would be to use an unbiased estimator that does not make use of the number of Type IV widgets in the warehouse or per box. Confirm that this estimate is `r format(N * mean(y), scientific = FALSE)` widgets. The other is to use a ratio estimator that uses the information about the number of Type IV widgets. Confirm that this estimate is `r format(taux * mean(y)/mean(x), scientific = FALSE)` widgets. 

```{r}
rm(list = ls())
```

