---
title: "Inclusion Probabilities in Design and Estimation"
output:
  html_document: 
    theme: readable
header-includes:
  - \usepackage{array}
  - \usepackage{booktabs}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, echo = FALSE, out.width = "100%", fig.align = "center", cache = FALSE, comment = "", dev = ifelse(knitr::is_html_output(), "png", "tikz"))
```

```{r packages, message = FALSE}
library(tidyverse)
library(kableExtra)
```

```{r utilities}
source("../../utilities.R")
```

`r ifelse(knitr::is_html_output(), paste("You can also download a [PDF](hw-inclusion.pdf) copy of this document.", sep = ""), "")`

### Computing Inclusion Probabilities

Suppose we have the following sampling design for a population of five elements.
\begin{align*}
  \mathcal{S}_1 & = \{\mathcal{E}_1,\mathcal{E}_3,\mathcal{E}_4\}, \ P(\mathcal{S}_1) = 0.1 \\ 
  \mathcal{S}_2 & = \{\mathcal{E}_1,\mathcal{E}_3,\mathcal{E}_5\}, \ P(\mathcal{S}_2) = 0.1 \\
  \mathcal{S}_3 & = \{\mathcal{E}_1,\mathcal{E}_4,\mathcal{E}_5\}, \ P(\mathcal{S}_3) = 0.1 \\ 
  \mathcal{S}_4 & = \{\mathcal{E}_2,\mathcal{E}_3,\mathcal{E}_4\}, \ P(\mathcal{S}_4) = 0.2 \\
  \mathcal{S}_5 & = \{\mathcal{E}_2,\mathcal{E}_3,\mathcal{E}_5\}, \ P(\mathcal{S}_5) = 0.2 \\ 
  \mathcal{S}_6 & = \{\mathcal{E}_2,\mathcal{E}_4,\mathcal{E}_5\}, \ P(\mathcal{S}_6) = 0.3 \\
\end{align*}
The table below shows the first-order ($\pi_i$) and second-order ($\pi_{ij}$) inclusion probabilities for the five elements in the population. 
```{r}
samp <- list(s1 = c(1,3,4), s2 = c(1,3,5), s3 = c(1,4,5), s4 = c(2,3,4), s5 = c(2,3,5), s6 = c(2,4,5))
prob <- c(0.1,0.1,0.1,0.2,0.2,0.3)

d <- matrix(NA, 5, 5)
for (i in 1:5) {
  for (j in 1:5) {
    d[i,j] <- 0
    for (k in 1:6) {
      if (i %in% samp[[k]] && j %in% samp[[k]]) {
        d[i,j] <- d[i,j] + prob[k]
      }
    }
  }
}

y <- c(1,2,4,2,1)

ht <- rep(NA,6)
hj <- rep(NA,6)
nt <- rep(NA,6)
for (i in 1:6) {
  ht[i] <- sum(y[samp[[i]]] / (diag(d)[samp[[i]]]))
  hj[i] <- ht[i] / sum(1 / (diag(d)[samp[[i]]]))
  nt[i] <- sum(1 / diag(d)[samp[[i]]])
}

d <- data.frame(cbind(1:5, d))
names(d) <- c("i", 1:5)
ktbl(d) %>% add_header_above(c(" " = 1, "j" = 5)) %>% column_spec(1, bold = TRUE)


```
Note that the values along the "diagonal" (i.e., from top left to bottom right) are the first-order inclusion probabilities, whereas the values not along this diagonal are the second-order inclusion probabilities. The reason why it makes sense to represent the inclusion probabilities this way is that the second-order inclusion probability of an element with itself (e.g., $\pi_{ii}$) equals the first-order inclusion probability (e.g., $\pi_i$). Verify these inclusion probabilities by computing them yourself from the sampling design. Note that the second-order inclusion probabilities are symmetric in the sense that $\pi_{ij} = \pi_{ji}$ so you only need to verify five first-order inclusion probabilities and ten second-order inclusion probabilities.

### Computing the Horvitz-Thompson Estimator

Consider the sampling design given in the previous problem. Assume that $y_1$ = `r y[1]`, $y_2$ = `r y[2]`, $y_3$ = `r y[3]`, $y_4$ = `r y[4]`, and $y_5$ = `r y[5]`. Confirm that the estimates of $\tau$ using the Horvitz-Thompson estimator are, approximately, `r vecprnt(round(ht,2))` for samples $\mathcal{S}_1$, $\mathcal{S}_2$, $\mathcal{S}_3$, $\mathcal{S}_4$, $\mathcal{S}_5$, and $\mathcal{S}_6$, respectively. Also use the Horvitz-Thompson estimator to compute an estimate of $\mu$ *two* ways: one assuming you know that the number of elements in the population is *known* to be five, and another where the number of elements in the population is *unknown*. You should confirm that the estimates based on the six samples with a known number of elements in the population are approximately `r vecprnt(round(ht/5, 2))`, respectively, and that the estimates based on the six samples with an unknown number of elements in the population are approximately `r vecprnt(round(ht/nt, 2))`, respectively.

### Generalized Horvitz-Thompson Estimator

```{r}
tx <- 90
p <- diag(as.matrix(d[,-1]))
x <- tx * p / 3
ty <- tx * sum(y[samp$s1]/p[samp$s1])/sum(x[samp$s1]/p[samp$s1])
```

Consider the sampling design described in the previous problems. Suppose that we have an auxiliary variable $x_i$ where $x_1$ = `r x[1]`, $x_2$ = `r x[2]`, $x_3$ = `r x[3]`, $x_4$ = `r x[4]`, and $x_5$ = `r x[5]`, and it is known that $\tau_x$ = `r tx`. Verify that the *generalized* Horvitz-Thompson estimator that is based on the *ratio estimator* produces an estimate of $\hat\tau_y$ $\approx$ `r round(ty,1)` if the sample is $\mathcal{S}_1$. 

### Sampling With Replacement

Consider the population of four elements, $\mathcal{P} = \{\mathcal{E}_1,\mathcal{E}_2,\mathcal{E}_3,\mathcal{E}_4\}$. Assume that the sample $\mathcal{S} = \{\mathcal{E}_1,\mathcal{E}_1,\mathcal{E}_4\}$ selected using sampling with replacement with given selection probabilities. The table below shows the value of the target variable for each element in the population and its selection probability. 
```{r}
d <- data.frame(Element = c("$\\mathcal{E}_1$", "$\\mathcal{E}_2$", "$\\mathcal{E}_3$", "$\\mathcal{E}_4$"), y = c(4,1,2,3), delta = c(0.1,0.2,0.3,0.4))
indx <- c(1,1,4)
n <- length(indx)
pi <- 1 - (1 - d$delta)^3
hh <- mean(d$y[indx]/d$delta[indx])
ht <- sum(d$y[unique(indx)]/pi[unique(indx)])
names(d)[2:3] <- c("$y_i$","$\\delta_i$")
ktbl(d)
```
Confirm that the estimate of $\tau$ using the Hansen-Hurwitz estimator `r rnd(hh,2)`, and that the estimate of $\tau$ using the Horvitz-Thompson estimator is `r rnd(ht,2)`.

### Line-Intercept Sampling

The figure below shows the the results of a survey using the line-intercept method. The selected objects, represented by circles, are those intercepted by the two transect lines. The value of the target variable is shown as the number within each object. 
```{r, fig.height = 4, fig.width = 9}
library(shape)

set.seed(102)

par(mai = c(0.1,0.1,0.1,0.1))

emptyplot(c(0, 100), c(0, 50), frame.plot = FALSE)

for (i in seq(0, 100, by = 5)) {
  lines(c(i,i), c(0,50), lty = 2, col = grey(0.8))
}
for (i in seq(0, 50, by = 5)) {
  lines(c(0, 100), c(i, i), lty = 2, col = grey(0.8))
}

rect <- function(mid, w) {
  x <- mid[1]
  y <- mid[2]
  lines(c(x - w/2, x + w/2), c(y - w/2, y - w/2))
  lines(c(x - w/2, x + w/2), c(y + w/2, y + w/2))
  lines(c(x - w/2, x - w/2), c(y - w/2, y + w/2))
  lines(c(x + w/2, x + w/2), c(y - w/2, y + w/2))
}

n.small <- 10
n.large <- 10

x.small <- rep(1, n.small)
y.small <- rep(1, n.small)
x.large <- rep(1, n.large)
y.large <- rep(1, n.large)

dcheck <- function(x.small, y.small, x.large, y.large, n.small, n.large) {
  d <- as.matrix(dist(cbind(c(x.small, x.large), c(y.small, y.large)), upper = TRUE))
  d.small <- d[1:n.small, 1:n.small]
  d.large <- d[(n.small + 1):(n.small + n.large), (n.small + 1):(n.small + n.large)]
  d.cross <- d[(n.small + 1):(n.small + n.large), 1:n.small]
  d.small <- min(d.small[lower.tri(d.small, diag = FALSE)])
  d.large <- min(d.large[lower.tri(d.large, diag = FALSE)])
  d.cross <- min(d.cross)
  return(any(d.small < 5, d.large < 10, d.cross < 10))
}

while (dcheck(x.small, y.small, x.large, y.large, n.small, n.large)) {
  x.small <- sample(seq(2.5, 97.5, by = 5), n.small, replace = TRUE)
  y.small <- sample(seq(2.5, 47.5, by = 5), n.small, replace = TRUE)
  x.large <- sample(seq(5, 95, by = 5), n.large, replace = TRUE)
  y.large <- sample(seq(5, 45, by = 5), n.large, replace = TRUE)
}

for (i in 1:n.small) {
  plotellipse(2.5, 2.5, c(x.small[i], y.small[i]), lwd = 1)
  text(x.small[i], y.small[i], rpois(1, 5) + 1)
}
for (i in 1:n.large) {
  plotellipse(5, 5, c(x.large[i], y.large[i]), lwd = 1)
  text(x.large[i], y.large[i], rpois(1, 10) + 1)
}

abline(v = c(11, 87))

y1 <- c(3,2,10)
y2 <- c(13,6)

d1 <- c(0.05,0.05,0.1)
d2 <- c(0.1,0.1)

ht1 <- sum(y1/d1)
ht2 <- sum(y2/d2)

y <- c(y1,y2)
p <- 1 - (1 - c(d1,d2))^2

ht.sep <- mean(c(ht1,ht2))
ht.all <- sum(y/p)
```
Confirm that the estimates of $\tau$ based on the left and right transect lines are `r ht1` and `r ht2`, respectively, using the Horvitz-Thompson estimator. These can be averaged together to produce and estimate of $\tau$ of `r ht.sep`. 

### Using Survey Weights

Suppose you have only the following information about a survey with an unknown design: the values of the target variable and the corresponding survey weights for the three elements in the sample.
```{r}
w <- c(40/10, 50/20, 60/30)
y <- c(5,3,2)
d <- data.frame(y,w)
tau <- sum(w * y)
names(d) <- c("Target Variable","Survey Weight")
ktbl(d)
```
Verify that based on this information alone an estimate of $\tau$ is $\hat\tau$ $=$ `r sum(w*y)`, and an estimate of $\mu$ is $\hat\mu$ $\approx$ `r round(sum(w*y)/sum(w), 1)`. 



