---
output:
  html_document: 
    theme: readable
  pdf_document: default
---

```{r, echo = FALSE, message = FALSE}
library(lubridate)
date <- "08-26-2022"
weekday <- wday(mdy(date), label = TRUE, abbr = FALSE)
month <- month(mdy(date), label = TRUE)
day <- day(mdy(date))
```

---
title: `r paste(weekday, ", ", month, " ", day, sep = "")`
output:
  html_document: 
    theme: readable
  pdf_document: default
header-includes:
  - \usepackage{float}
  - \usepackage{booktabs}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, out.width = "100%", fig.align = "center", cache = FALSE, dev = ifelse(knitr::is_html_output(), "png", "pdf"))
```

```{r packages}
library(tidyverse)
suppressWarnings(library(kableExtra))
```

```{r utilities}
source("../../utilities.R")
```

`r ifelse(knitr::is_html_output(), paste("You can also download a [PDF](lecture-", date, ".pdf) copy of this lecture.", sep = ""), "")`

## Sampling Distributions --- Continued

```{r}
pop <- c(1,2,2,7)

mu1 <- mean(pop)
sigma1 <- var(pop)

N <- length(pop)
n <- 2

m <- apply(combn(pop, n), 2, mean)
t <- apply(combn(pop, n), 2, function(z) N/n * sum(z))

elem <- paste("\\mathcal{E}_{", 1:N, "}", sep = "")
space <- paste("$\\{", apply(combn(elem, n), 2, function(z) paste(z, collapse = ",")), "\\}$", sep = "")
```

**Example**: Suppose we have a population of $N$ = `r N` elements where the target variable values are `r vecprnt(pop)`. The table below shows the **sample space** for a simple random sampling design with $n$ = `r n` as well as the values of the estimators. $\hat\tau$ and $\bar{y}$ for each sample.
```{r}
d <- data.frame(Sample = space, tau = t, m = m)
names(d) <- c("Sample", "$\\hat\\tau$", "$\\bar{y}$")
ktbl(d)
```
Note that a **sample space** is simply the set of all possible samples. Here there are `r choose(N,n)` samples in the sample space. The table below shows the *sampling distributions* of $\bar{y}$ and $\hat\tau$. 
```{r}
d <- data.frame(m = m, t = t) %>% group_by(m,t) %>% 
  summarize(f = n()) %>% ungroup() %>% mutate(p = f/sum(f)) %>% select(m, t, p)
e1 <- with(d, sum(m * p))
v1 <- with(d, sum((m - e1)^2 * p))
dt <- d
dt$p <- as.character(MASS::fractions(dt$p))
names(dt) <- c("$\\bar{y}$","$\\hat\\tau$","Probability")
ktbl(dt)
```
The figures below show the sampling distributions of $\bar{y}$ and $\hat\tau$.
```{r, fig.height = 3}
p <- ggplot(d, aes(x = m, y = p)) + theme_minimal()
p <- p + geom_segment(aes(xend = m, yend = 0), alpha = 0.5) + geom_point()
p <- p + labs(x = tex("$\\bar{y}$"), y = "Probability")
p1 <- p
p <- ggplot(d, aes(x = t, y = p)) + theme_minimal()
p <- p + geom_segment(aes(xend = t, yend = 0), alpha = 0.5) + geom_point()
p <- p + labs(x = tex("$\\hat{\\tau}$"), y = "Probability")
p2 <- p
cowplot::plot_grid(p1, p2)
```

```{r}
set.seed(123)

pop <- sort(sample(1:10, 10, replace = TRUE))

mu2 <- mean(pop)
sigma2 <- var(pop)

N <- length(pop)
n <- 4

m <- apply(combn(pop, n), 2, mean)
t <- apply(combn(pop, n), 2, function(z) N/n * sum(z))

elem <- paste("\\mathcal{E}_{", 1:N, "}", sep = "")
space <- paste("$\\{", apply(combn(elem, n), 2, function(z) paste(z, collapse = ",")), "\\}$", sep = "")
```

**Example**: Suppose now we have a population of $N$ = `r N` elements where the target variable values are `r vecprnt(pop)`. The table below shows the (abbreviated) **sample space** for a simple random sampling design with $n$ = `r n` as well as the values of the estimators $\hat\tau$ and $\bar{y}$ for each sample.
```{r}
d <- data.frame(Sample = space, tau = t, m = m)
names(d) <- c("Sample", "$\\hat\\tau$", "$\\bar{y}$")
ktbl(headtail(d, 8))
```
Here there are `r choose(N,n)` samples in the sample space. The table below shows the *sampling distributions* of $\bar{y}$ and $\hat\tau$. 
```{r}
d <- data.frame(m = m, t = t) %>% group_by(m,t) %>% 
  summarize(f = n()) %>% ungroup() %>% mutate(p = f/sum(f)) %>% select(m, t, p)
e2 <- with(d, sum(m * p))
v2 <- with(d, sum((m - e1)^2 * p))
dt <- d
names(dt) <- c("$\\bar{y}$","$\\hat\\tau$","Probability")
ktbl(dt)
```
The figures below show the sampling distributions of $\bar{y}$ and $\hat\tau$.
```{r, fig.height = 3}
p <- ggplot(d, aes(x = m, y = p)) + theme_minimal()
p <- p + geom_segment(aes(xend = m, yend = 0), alpha = 0.5) + geom_point()
p <- p + labs(x = tex("$\\bar{y}$"), y = "Probability")
p1 <- p
p <- ggplot(d, aes(x = t, y = p)) + theme_minimal()
p <- p + geom_segment(aes(xend = t, yend = 0), alpha = 0.5) + geom_point()
p <- p + labs(x = tex("$\\hat{\\tau}$"), y = "Probability")
p2 <- p
cowplot::plot_grid(p1, p2)
```

## Means and Variances of Random Variables

If we have a *discrete* random variable $X$, then the **mean** of that random variable (also called its *expectation* or *expected value*) is 
$$
  E(X) = \sum_x xP(x),
$$
where the $x$ below $\sum$ indicates that we sum over all values of $x$. 

**Example**: We can easily confirm that based on the sampling distributions for the smaller population above that $E(\bar{y}) =$ `r e1` and $E(\hat\tau) =$ `r e1*4`. 

\vspace{7cm}

But there is a shortcut if we use simple random sampling. For *any* simple random sampling design, *it can be shown that* $E(\bar{y}) = \mu$ and $E(\hat\tau) = \tau$. When the mean of an estimator equals what is being estimated, then we say that the estimator is an **unbiased** estimator (otherwise the estimator is a **biased** estimator). 

**Warning**: We have discussed *two* distinct means here: the population mean ($\mu$), and the mean of an estimator --- e.g., $E(\bar{y})$ and $E(\hat\tau)$. Don't confuse them!

\pagebreak

If we have a discrete random variable $X$, then the **variance** of that random variable is 
$$
  \text{V}(X) = \sum_x [x - E(X)]^2P(x).
$$
**Example**: We can easily confirm that based on the sampling distributions for the smaller population above that $\text{V}(\bar{y}) \approx$ `r round(v1,2)` and $\text{V}(\hat\tau) \approx$ `r round(v1*4^2,2)`.

\vspace{5cm}

It can be shown that *under simple random sampling* that the variances can be computed as
$$
  \text{V}(\bar{y}) = \left(1 - \frac{n}{N}\right)\frac{\sigma^2}{n} \ \ \ \text{and} \ \ \ 
  \text{V}(\hat\tau) = N^2\left(1 - \frac{n}{N}\right)\frac{\sigma^2}{n},
$$
where $\sigma^2$ is the *population variance*
$$
  \sigma^2 = \frac{1}{N-1}\sum_{i=1}^N(y_i - \mu)^2, 
$$
and where $\mu$ is the population mean.[^denominator]

[^denominator]: Some textbooks will define $\sigma^2$ by dividing by $N$ instead of $N-1$. This can be done but it changes the formulas for $V(\bar{y})$ and $V(\hat\tau)$ in a way that I find awkward for later developments. In terms of our interpretation of $\sigma^2$ this is inconsequential since we rarely care about $\sigma^2$ in isolation and the difference between $N$ and $N-1$ is very small when $N$ is large which is frequently the case in survey sampling. 

**Example**: For the smaller population above $\sigma^2 =$ `r round(sigma1,2)`, and for the larger population $\sigma^2 \approx$ `r round(sigma2,2)`. What are $\text{V}(\bar{y})$ and $\text{V}(\hat\tau)$ for each design?

\vspace{5cm}

**Warning**: We've discussed *two* distinct variances here: the *population variance* ($\sigma^2$) and the *variance of an estimator* --- e.g., $\text{V}(\bar{y})$ and $\text{V}(\hat\tau)$. Don't confuse them!

\pagebreak

How do $n$ and $N$ affect the variance of these estimators?
```{r, fig.height = 3}
d <- expand.grid(N = c(10,20,40,80,160), n = 1:160) %>% filter(n <= N) %>%
  mutate(var = (1 - n/N)/n)

p <- ggplot(d, aes(x = n, y = var, color = factor(N))) + theme_minimal()
p <- p + geom_line() + scale_y_continuous(breaks = c(0, 0.25, 0.5, 0.75, 1), 
  minor_breaks = NULL, labels = c("","","","",tex("$\\sigma^2$")))
p <- p + labs(x = tex("$n$"), color = tex("$N$"), y = tex("$V(\\bar{y})$"))
p <- p + scale_x_continuous(breaks = c(1, seq(10, 160, by = 10)))
plot(p)
```

## Central Limit Theorem

What can we say about the *shape* of the sampling distribution of $\bar{y}$ or $\hat\tau$? 
The **central limit theorem** for simple random sampling states that as $n$, $N$, and $N-n$ increase, and assuming some other rather technical but usually applicable conditions, the sampling distribution of $\bar{y}$ "approaches" a normal distribution. This implies the same behavior for the sampling distribution of $\hat\tau$.

**Example**: Consider simple random sampling design with a population of $N$ = 50 elements and a sample size of $n$ = 5. The values of the target variable in the population were selected randomly from the integers from 1 and 100. The figure below shows the sampling distribution of $\bar{y}$.
```{r, cache = TRUE, fig.height = 3}
set.seed(123)
pop <- sample(1:100, 50, replace = TRUE)

N <- length(pop)
n <- 5

m <- apply(combn(pop, n), 2, mean)

d <- data.frame(m = m) %>% group_by(m) %>% 
  summarize(f = n()) %>% mutate(p = f/sum(f)) %>% select(m, p)

p <- ggplot(d, aes(x = m, y = p)) + theme_minimal()
p <- p + geom_segment(aes(xend = m, yend = 0), alpha = 0.5)
p <- p + labs(x = tex("$\\bar{y}$"), y = "Probability")
plot(p)
```

