---
title: "Lectures"
output: 
  html_document: 
    theme: readable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r, message = FALSE}
library(lubridate)
d <- dmy("25-08-2025","27-08-2024","29-08-2025") + rep(7 * c(0:17), each = 3)
d <- paste(wday(d, label = TRUE, abbr = TRUE), ", ", month(d, label = TRUE), " ", day(d), ".", sep = "")
i <- 0
```

&nbsp;

`r i <- i + 1; d[i]` [Introduction, fundamental concepts including elements, sampling units, populations, samples, sampling designs](lecture-08-25-2025.html).

`r i <- i + 1; d[i]` [Introduction to simple random sampling, parameters and estimators, introduction to sampling distributions](lecture-08-27-2025.html).

`r i <- i + 1; d[i]` [Sampling distributions --- continued, sampling distributions $\bar{y}$ and $\hat\tau$ under SRS](lecture-08-29-2025.html). A [homework assignment on simple random sampling](hw-srs.html) is now available. After today's lecture you should be able to do the first section titled "Sampling Distributions". 
 
***

`r i <- i + 1; d[i]` Holiday.

`r i <- i + 1; d[i]` [Sampling with replacement, central limit theorem, interpreting normal sampling distributions, standard errors and bounds on the error of estimation, confidence intervals](lecture-09-03-2025.html).

`r i <- i + 1; d[i]` [Simulation studies, categorical target variables, domain estimators](lecture-09-05-2025.html).
 
***

`r i <- i + 1; d[i]` [Sample size selection, summary of inference in the context of simple random sampling](lecture-09-08-2025.html).

`r i <- i + 1; d[i]` [Introduction to stratified random sampling](lecture-09-10-2025.html).

`r i <- i + 1; d[i]` [Optimum allocation](lecture-09-12-2025.html). A set of [homework problems](hw-stratified.html) concerning stratified sampling and related topics is now available. The [study guide](e1-studyguide.html) for the first examination is also now available.
 
***

`r i <- i + 1; d[i]` [Optimum allocation --- continued, design effect, effective sample size](lecture-09-15-2025.html).

`r i <- i + 1; d[i]` [Optimum stratification, stratified random sampling via double sampling](lecture-09-17-2025.html).

`r i <- i + 1; d[i]` [Post-stratification, survey weights and re-weighting](lecture-09-19-2025.html).
 
***

`r i <- i + 1; d[i]` [Inferences concerning stratum parameters](lecture-09-22-2025.html).

`r i <- i + 1; d[i]` [Using the `survey` package in R for inferences from simple and stratified random sampling designs](lecture-09-24-2025.html).

`r i <- i + 1; d[i]` **First examination**. 
 
***

`r i <- i + 1; d[i]` [Estimating ratios](lecture-09-29-2025.html).  

`r i <- i + 1; d[i]` [Introduction to ratio estimators](lecture-10-01-2025.html).

`r i <- i + 1; d[i]` [Ratio estimators --- continued](lecture-10-03-2025.html).
 
***

`r i <- i + 1; d[i]` [Mean squared error, double sampling for ratio estimators, ratio estimators for stratified random sampling designs](lecture-10-06-2025.html). A set of [homework questions](hw-ratio.html) is now available concerning ratio estimators, regression estimators, and re-weighting and calibration. 

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 
 
***

`r i <- i + 1; d[i]` **Second examination**.

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 
 
`r i <- i + 1; d[i]` 
 
***

`r i <- i + 1; d[i]` Fall recess.

`r i <- i + 1; d[i]` Fall recess.

`r i <- i + 1; d[i]` Fall recess.
 
***

`r i <- i + 1; d[i]` 
 
`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` **Third examination**.
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 

***

Monday, Dec 15. **Final examination**, 12:45 to 2:45 PM.


