---
title: "Lectures"
output: 
  html_document: 
    theme: readable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r, message = FALSE}
library(lubridate)
d <- dmy("22-08-2022","24-08-2022","26-08-2022") + rep(7 * c(0:17), each = 3)
d <- paste(wday(d, label = TRUE, abbr = TRUE), ", ", month(d, label = TRUE), " ", day(d), ".", sep = "")
i <- 0
```

&nbsp;

`r i <- i + 1; d[i]` [Fundamental definitions: elements, sampling units, populations, samples, sampling designs](lecture-08-22-2022.html).

`r i <- i + 1; d[i]` [Simple random sampling, population totals and means, and an introduction to sampling distributions](lecture-08-24-2022.html).

`r i <- i + 1; d[i]` [Sampling distributions of $\bar{y}$ and $\hat\tau$ under simple random sampling designs](lecture-08-26-2022.html). A set of [homework problems concerning simple random sampling](hw-srs.html) is now available. You should be able to start on the problems in the section titled "Sampling Distributions" after today's lecture. Future lectures will cover domain estimation and sample size selection. 
 
***

`r i <- i + 1; d[i]` [Sampling with replacement, the central limit theorem, standard errors, confidence intervals](lecture-08-29-2022.html). 

`r i <- i + 1; d[i]` [Simulation studies, categorical target variables, domain estimators](lecture-08-31-2022.html).

`r i <- i + 1; d[i]` [Sample size selection, summary of results for simple random sampling](lecture-09-02-2022.html).
 
***

`r i <- i + 1; d[i]` Labor Day.

`r i <- i + 1; d[i]` [Introduction to stratified random sampling](lecture-09-07-2022.html). A set of [homework problems](hw-stratified.html) on stratified random sampling is now available. The [study guide](e1-studyguide.html) for the first examination is also now available.

`r i <- i + 1; d[i]` [Optimum allocation for stratified random sampling](lecture-09-09-2022.html). 
 
***

`r i <- i + 1; d[i]` [Effectiveness of optimum allocation, stratified versus simple random sampling, design effect and effective sample size](lecture-09-12-2022.html).

`r i <- i + 1; d[i]` [Optimum stratification, double sampling](lecture-09-14-2022.html).

`r i <- i + 1; d[i]` [Post-stratification, survey weights](lecture-09-16-2022.html).
 
***

`r i <- i + 1; d[i]` [Inferences concerning strata parameters](lecture-09-19-2022.html). Note: I added one additional item to the study guide for this material. 

`r i <- i + 1; d[i]` [Using the **survey** package in R](lecture-09-21-2022.html).

`r i <- i + 1; d[i]` **First examination**. 
 
***

`r i <- i + 1; d[i]` [Estimation of a ratio of totals](lecture-09-26-2022.html).

`r i <- i + 1; d[i]` [Introduction to ratio estimators](lecture-09-28-2022.html).

`r i <- i + 1; d[i]` [Domain total estimators revisited, the geometry and performance of ratio estimators](lecture-09-30-2022.html). A set of [homework problems](hw-ratio.html) on estimation ratios, ratio estimators, and related topics is now available.
 
***

`r i <- i + 1; d[i]` [Bias and mean squared error, double sampling for ratio estimators, ratio estimators for stratified random sampling designs](lecture-10-03-2022.html).

`r i <- i + 1; d[i]` [Regression estimators, the prediction perspective](lecture-10-05-2022.html).

`r i <- i + 1; d[i]` [Reweighting, calibration](lecture-10-07-2022.html).
 
***

`r i <- i + 1; d[i]` [Introduction to (one-stage) cluster sampling](lecture-10-10-2022.html). The [study guide](e2-studyguide.html) for the next examination is now available. 

`r i <- i + 1; d[i]` [Cluster sampling --- continued, estimators of $\tau$, the design effect of cluster sampling designs](lecture-10-12-2022.html). A [set of homework problems](hw-cluster.html) on cluster sampling is now available.

`r i <- i + 1; d[i]` [The Hansen-Hurwitz estimator when sampling with replacement, PPS cluster sampling](lecture-10-14-2022.html). 
 
***

`r i <- i + 1; d[i]` [Stratified cluster sampling designs, systematic sampling](lecture-10-17-2022.html).

`r i <- i + 1; d[i]` [Systematic sampling --- continued, ranked set sampling](lecture-10-19-2022.html). 

`r i <- i + 1; d[i]` [Inclusion probabilities --- revisited, themes in survey sampling](lecture-10-21-2022.html).
 
***

`r i <- i + 1; d[i]` **Second examination**.

`r i <- i + 1; d[i]` [Two-stage cluster sampling](lecture-10-26-2022.html).

`r i <- i + 1; d[i]` [Optimum sample sizes for two-stage cluster sampling, multi-stage cluster sampling](lecture-10-28-2022.html). A [set of homework problems](hw-twostage.html) on two-stage cluster sampling is now available.
 
***

`r i <- i + 1; d[i]` [Introduction to abundance estimation](lecture-10-31-2022.html).

`r i <- i + 1; d[i]` [Abundance estimation with diagnostic tests, introduction to mark-recapture designs](lecture-11-02-2022.html).

`r i <- i + 1; d[i]` 
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]`
 
`r i <- i + 1; d[i]` 
 
***

`r i <- i + 1; d[i]` Fall recess.

`r i <- i + 1; d[i]` Fall recess. 

`r i <- i + 1; d[i]` Fall recess. 
 
***

`r i <- i + 1; d[i]`

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` **Third examination**. 
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 

***

Wednesday, Dec 14. **Final examination**, 12:45 to 2:45 PM.


