---
output:
  html_document: 
    theme: readable
  pdf_document: default
---

```{r, echo = FALSE, message = FALSE}
library(lubridate)
x <- strsplit(this.path::this.path(), split = "/")[[1]]
x <- strsplit(x[length(x)], split = "-")[[1]]
date <- paste(x[2], x[3], year(Sys.Date()), sep = "-")
weekday <- wday(mdy(date), label = TRUE, abbr = FALSE)
month <- month(mdy(date), label = TRUE, abbr = FALSE)
day <- day(mdy(date))
```

---
title: `r paste(weekday, ", ", month, " ", day, sep = "")`
output:
  html_document: 
    theme: readable
  pdf_document: default
urlcolor: blue
header-includes:
  - \usepackage{float}
  - \usepackage{booktabs}
  - \usepackage{array}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, out.width = "100%", fig.align = "center", cache = FALSE, dev = ifelse(knitr::is_html_output(), "png", "pdf"))
```

```{r packages}
library(tidyverse)
suppressWarnings(library(kableExtra))
```

`r ifelse(knitr::is_html_output(), paste("You can also download a [PDF](lecture-", date, ".pdf) copy of this lecture.", sep = ""), "")`

```{r}
ktbl <- function(d, caption = NULL, align = "c") {
  d %>% kable(format = ifelse(knitr::is_html_output(), "html", "latex"), 
    booktabs = TRUE, linesep = "", align = align, escape = FALSE, caption = caption) %>%
    kable_styling(full_width = FALSE, bootstrap_options = "hover", latex_options = "hold_position")
}

headtail <- function(d, n) {
  for (i in 1:ncol(d)) {
    d[,i] <- as.character(d[,i])
  }
  d1 <- d[1:n,]
  d2 <- rep("$\\vdots$", ncol(d))
  d3 <- d[nrow(d),]
  y <- rbind(d1, d2, d3)
  rownames(y) <- NULL
  return(y)
}

tex <- function(x, html = knitr::opts_chunk$get("dev") %in% c("png","pdf"), ...) {
  ifelse(html, latex2exp::TeX(x, ...), x)
}

noyaxis <- theme(axis.title.y = element_blank(), axis.text.y = element_blank(), 
  axis.ticks.y = element_blank(), axis.line.y = element_blank())

rtrunc <- function(f, n, a, b, ...) {
  y <- rep(NA,n)
  for (i in 1:n) {
    repeat{
      y[i] <- f(n = 1, ...)
      if (a <= y[i] & y[i] <= b) {
        break
      }
    }
  }
  return(y)
}
```


## Inclusion Probabilities --- Revisited

Recall that an *inclusion probability* ($\pi_i$) of the $i$-th element is the probability that the element will be in the selected sample.

For *simple random sampling*, all elements have an inclusion probability of $\pi_i = n/N$.

For *stratified random sampling*, the inclusion probability of the $i$-th element is $\pi_i = n_j/N_j$ if the element is in stratum $j$. 

What are the inclusion probabilities for a *cluster sampling design*? It depends on how the *clusters* are sampled. The inclusion probability of an element is the same as that of its cluster. Consider three ways to sample clusters.

1. Simple random sampling.

\vspace{1cm}

2. Stratified random sampling. 

\vspace{1cm}

3. (Selection) probability proportional to (cluster) size (with replacement). 

\vspace{1cm}

Question: Some people [define](https://www.investopedia.com/terms/s/simple-random-sample.asp) simple random sampling as one where every element has the same inclusion probability. Is this correct? Why or why not? 

\pagebreak

## Some Themes in Survey Sampling

### Auxiliary Variables

An *auxiliary variable* is a variable that can be observed for every element (or sometimes the sampling unit) in the population.

Uses of auxiliary variables:

1. Improving a *sampling design*. What examples of this have we seen so far?

0. Improving an *estimator*. What examples of this have we seen so far? 

### Double Sampling

A *double sampling* can be used to facilitate the use of an auxiliary variable in compute an estimator.

1. How might we use double sampling to facilitate the use of estimators used in *stratified random sampling*?

0. How might we use double sampling to facilitate the use of a *ratio or regression estimator*?

### Simple Random Sampling in Complex Sampling Designs

Some complex sampling designs using simple random sampling in some way.

1. How might *stratified random sampling* use simple random sampling?

0. How might *cluster sampling* use simple random sampling?

### Sampling Design Decisions

Fundamentally, design decisions involve specifying the sample space of possible samples, and sometimes also changing their probabilities.

1. What design decisions are made in *simple random sampling*?

0. What design decisions are made in *stratified random sampling*?

0. What design decisions are made in *cluster sampling*?

### Relative Precision of the Designs

1. How does *stratified random sampling* (usually) compare to simple random sampling?

0. How does *cluster sampling* compare (usually) to simple random sampling? 



