---
title: "Lectures"
output: 
  html_document: 
    theme: readable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r, message = FALSE}
library(lubridate)
d <- dmy("21-08-2023","23-08-2023","25-08-2023") + rep(7 * c(0:17), each = 3)
d <- paste(wday(d, label = TRUE, abbr = TRUE), ", ", month(d, label = TRUE), " ", day(d), ".", sep = "")
i <- 0
```

&nbsp;

`r i <- i + 1; d[i]` [Introduction, fundamental concepts: elements, sampling units, populations, samples, sampling designs](lecture-08-21-2023.html).

`r i <- i + 1; d[i]` [Introduction to simple random sampling designs](lecture-08-23-2023.html). 

`r i <- i + 1; d[i]` [Sampling distributions of $\bar{y}$ and $\hat\tau$ under simple random sampling designs](lecture-08-25-2023.html). A set of [homework problems concerning simple random sampling](hw-srs.html) is now available. You should be able to start on the problems in the section titled "Sampling Distributions" after today's lecture. Future lectures will cover the other topics in that homework set.
 
***

`r i <- i + 1; d[i]` [Sampling with replacement, the central limit theorem, standard errors, bounds on the error of estimation, and confidence intervals](lecture-08-28-2023.html).

`r i <- i + 1; d[i]` [Simulation studies, categorical target variables, domain estimators](lecture-08-30-2023.html).

`r i <- i + 1; d[i]` [Sample size selection, overview of simple random sampling](lecture-09-01-2023.html).
 
***

`r i <- i + 1; d[i]` Labor Day.

`r i <- i + 1; d[i]` [Introduction to stratified random sampling](lecture-09-06-2023.html). A set of [homework problems](hw-stratified.html) on stratified random sampling is now available. The [study guide](e1-studyguide.html) for the first examination is also now available.

`r i <- i + 1; d[i]` [Optimum allocation for stratified random sampling](lecture-09-08-2023.html).
 
***

`r i <- i + 1; d[i]` [Effectiveness of optimum allocation, stratified versus simple random sampling, design effect and effective sample size](lecture-09-11-2023.html).

`r i <- i + 1; d[i]` [Optimum stratification, double sampling to use stratification](lecture-09-13-2023.html).

`r i <- i + 1; d[i]` [Post-stratification, survey weights, and re-weighting](lecture-09-15-2023.html).
 
***

`r i <- i + 1; d[i]` [Inferences concerning stratum parameters](lecture-09-18-2023.html).

`r i <- i + 1; d[i]` [Survey sampling inferences using the **survey** package in R](lecture-09-20-2023.html).

`r i <- i + 1; d[i]` **First examination**. 
 
***

`r i <- i + 1; d[i]` [Estimation of a ratio of totals](lecture-09-25-2023.html).

`r i <- i + 1; d[i]` [Two estimators of a ratio of totals, ratio estimators, performance and advantages of ratio estimators](lecture-09-27-2023.html).

`r i <- i + 1; d[i]` [Two estimators of a domain total revisited, geometry and performance of ratio estimators, sources of auxiliary variables for ratio estimator](lecture-09-29-2023.html).
 
***

`r i <- i + 1; d[i]` [Bias and mean squared error for ratio estimators, double sampling for ratio estimators, ratio estimators for stratified random sampling designs](lecture-10-02-2023.html). A set of [homework problems](hw-ratio.html) on ratio estimators and related topics is now available.  

`r i <- i + 1; d[i]` [Regression estimators, the prediction perspective](lecture-10-04-2023.html).

`r i <- i + 1; d[i]` [Re-weighting revisited, calibration, raking](lecture-10-06-2023.html). The [study guide](e2-studyguide.html) for the second examination is now available. 
 
***

`r i <- i + 1; d[i]` [Introduction to (one-stage) cluster sampling](lecture-10-09-2023.html). 

`r i <- i + 1; d[i]` [Estimators of $\tau$ for cluster sampling designs, the design effect of cluster sampling](lecture-10-11-2023.html). A set of [homework problems](hw-cluster.html) on cluster sampling is now available. 

`r i <- i + 1; d[i]` [Sampling with replacement with unequal selection probabilities, the Hansen-Hurwitz estimator, cluster sampling with probabilities proportional to size](lecture-10-13-2023.html).
 
***

`r i <- i + 1; d[i]` [Stratified sampling, introduction to systematic sampling](lecture-10-16-2023.html).

`r i <- i + 1; d[i]` [Stratified sampling --- continued, ranked set sampling](lecture-10-18-2023.html).

`r i <- i + 1; d[i]` 
 
***

`r i <- i + 1; d[i]` **Second examination**.

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]`
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 
 
`r i <- i + 1; d[i]` 
 
***

`r i <- i + 1; d[i]` Fall recess.

`r i <- i + 1; d[i]` Fall recess. 

`r i <- i + 1; d[i]` Fall recess. 
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` **Third examination**. 
 
***

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 

`r i <- i + 1; d[i]` 

***

Thursday, Dec 14. **Final examination**, 12:45 to 2:45 PM.


