---
output:
  html_document: 
    theme: readable
  pdf_document: default
---

```{r, echo = FALSE, message = FALSE}
library(lubridate)
date <- "08-19-2024"
weekday <- wday(mdy(date), label = TRUE, abbr = FALSE)
month <- month(mdy(date), label = TRUE)
day <- day(mdy(date))
```

---
title: `r paste(weekday, ", ", month, " ", day, sep = "")`
output:
  html_document: 
    theme: readable
  pdf_document: default
urlcolor: blue
header-includes:
  - \usepackage{float}
  - \usepackage{booktabs}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, out.width = "100%", fig.align = "center", cache = FALSE, dev = ifelse(knitr::is_html_output(), "png", "pdf"))
```

```{r packages}
library(tidyverse)
suppressWarnings(library(kableExtra))
```

```{r utilities}
source("../../utilities.R")
```

`r ifelse(knitr::is_html_output(), paste("You can also download a [PDF](lecture-", date, ".pdf) copy of this lecture.", sep = ""), "")`

## Introduction

What this class *is* about:

1. *Sampling design*. How can we obtain a "useful" sample of things from a population of things? How can we do this effectively while also controlling survey cost? How can we use outside information to create a design that might produce a more useful sample?

2. *Inference*. How can we characterize the "usefulness" of a particular sampling design? How do we use the information in the sample? How can we use information outside the sample to improve inferences? 

This class is *not* about *measurement* (e.g., questionnaire design, disease diagnosis, quantification of foliage cover), although that is *very* important. 

Example applications:

How many Hobbits living in the Shire have foot lice? (epidemiology)

What is the average lead content of pottery shards at a site? (archaeology)

What is the average algebra skills test score at a high school? (education)

What proportion of undergraduates at UI use an Android phone? (marketing)

How many otter dens are there along the cost of Scotland? (ecology)

## Advantages and Disadvantages of Sampling

What are the advantages and disadvantages relative to a complete census? 

*Advantages*:

1. reduced cost
2. faster
3. greater scope
4. greater accuracy

*Disadvantages*:

1. loss of accuracy
2. requires technical expertise 

## Elements, Sampling Units, Samples, and Populations

An **element** is the fundamental *observational unit* (i.e., the "thing" from which we get a value of a variable of interest). Let $\mathcal{E}_i$ denote the $i$-th element. For each element we can observe the value of a *target variable* ($y_i$).

**Example**: Suppose we have 12 elements: $\mathcal{E}_1, \mathcal{E}_2, \dots, \mathcal{E}_{12}$. 

A **population** is the set of *all* elements of interest. We could denote this as $\mathcal{P}$. 

**Example**: The population is $\mathcal{P} = \left\{ \mathcal{E}_1, \mathcal{E}_2, \dots, \mathcal{E}_{12} \right\}$.

A **sampling unit** is a set of *one or more elements* such that if one of the elements within the sampling unit are in a sample, then all of the units within that sampling unit are in the sample (i.e., the elements in a sampling unit are always included in a sample *together*). The sampling units partition the population of units in such a way that each element appears in one and only one sampling unit. The $i$-th sampling unit will be denoted as $\mathcal{U}_i$.

**Example**: The sampling units *could be* as follows:
\begin{align*}
  \mathcal{U}_1 & = \left\{\mathcal{E}_1,\mathcal{E}_2\right\} \\
  \mathcal{U}_2 & = \left\{\mathcal{E}_3\right\} \\
  \mathcal{U}_3 & = \left\{\mathcal{E}_4, \mathcal{E}_5, \dots, \mathcal{E}_9\right\} \\
  \mathcal{U}_4 & = \left\{\mathcal{E}_{10}, \mathcal{E}_{11}, \mathcal{E}_{12} \right\}
\end{align*}

A **sampling frame** is a list (literal or conceptual) of all of the *sampling units* in the population.

A **sample** is a set of sampling units, and thus also a subset of the elements in a population. These are the only elements for which we observe the value of the target variable. Samples will be denoted as $\mathcal{S}$ or $\mathcal{S}_i$ if we need to refer to more than one sample.

**Example**: The sample
$$
  \mathcal{S} = \left\{\mathcal{U}_1, \mathcal{U}_4\right\} = \left\{\mathcal{E}_1, \mathcal{E}_2, \mathcal{E}_{10}, \mathcal{E}_{11}, \mathcal{E}_{12} \right\}
$$
contains two sampling units and five elements. The sample 
$$  
  \mathcal{S} = \left\{\mathcal{U}_1, \mathcal{U}_2\right\} = \left\{\mathcal{E}_1, \mathcal{E}_2, \mathcal{E}_3 \right\}
$$
contains two sampling units and three elements. 

## Sampling Designs

A (probability) **sampling design** is (a) the list of all possible samples and (b) the probability of each possible sample. 

**Example**: Here is a sampling design for the population $\mathcal{P} = \left\{ \mathcal{E}_1, \mathcal{E}_2, \dots, \mathcal{E}_{12} \right\}$ based on the sampling units defined earlier.
\begin{align*}
\mathcal{S}_1 & = \left\{\mathcal{U}_1,\mathcal{U}_2,\mathcal{U}_3\right\} = \left\{\mathcal{E}_1,\mathcal{E}_2,\mathcal{E}_3,\mathcal{E}_4,\mathcal{E}_5,\mathcal{E}_6,\mathcal{E}_7,\mathcal{E}_8,\mathcal{E}_9 \right\} \\
\mathcal{S}_2 & = \left\{\mathcal{U}_1,\mathcal{U}_2,\mathcal{U}_4\right\} = \left\{\mathcal{E}_1,\mathcal{E}_2,\mathcal{E}_3,\mathcal{E}_{10}, \mathcal{E}_{11}, \mathcal{E}_{12} \right\} \\
\mathcal{S}_3 & = \left\{\mathcal{U}_1,\mathcal{U}_3,\mathcal{U}_4\right\} = 
\left\{\mathcal{E}_1,\mathcal{E}_2,\mathcal{E}_4,\mathcal{E}_5,\mathcal{E}_6,\mathcal{E}_7,\mathcal{E}_8,\mathcal{E}_9,\mathcal{E}_{10}, \mathcal{E}_{11}, \mathcal{E}_{12} \right\} \\
\mathcal{S}_4 & = \left\{\mathcal{U}_2,\mathcal{U}_3,\mathcal{U}_4\right\} = 
\left\{\mathcal{E}_3,\mathcal{E}_4,\mathcal{E}_5,\mathcal{E}_6,\mathcal{E}_7,\mathcal{E}_8,\mathcal{E}_9,\mathcal{E}_{10}, \mathcal{E}_{11}, \mathcal{E}_{12}\right\}
\end{align*}
\begin{align*}
P(\mathcal{S}_1) & = 0.2 \\
P(\mathcal{S}_2) & = 0.3 \\
P(\mathcal{S}_3) & = 0.4 \\
P(\mathcal{S}_4) & = 0.1 \\
\end{align*}

## Probability Sampling

Why *probability sampling*? 

1. To help avoid intentional or unintentional biases in our inferences from the sample to the population. 

2. Probability is mathematically tractable --- we may not know what the sample will be, but we can speak volumes to how certain (or uncertain) we can be about what it *might* be.

 




